%Spelling checked 2004-04-06

The following four documents are designed to be templates to modify as desired.
 
The first one, the \textbf{Generic Vision Statement}, is a pithy statement of the group's purpose and values. Ideally, this statement will change only slightly over time. Therefore, the statement should, by design, be broad and general.
 
The second, \textbf{Generic Principles}, is related to the first in that in this document, each idea in the Vision Statement is defined and described at length and in great detail. It is not broad and general; it \emph{is} the definition of the specific principles and purposes adopted by the group's consent. This document is designed to change frequently; theoretically, with every decision. It is where decisions made by the group are accumulated according to the purpose or value it refines.
 
The third, \textbf{Generic Nonprofit Bylaws}, is the bylaws for nonprofit organizations using consensus decisionmaking. In theory, organizations that use consensus would be horizontally structured. This means there would be no “Board of Directors” who legally would have more power and responsibility than the rest of the membership. A structure that utilizes a Board of Directors is called an oligarchy. There is a “Board of Directors” in this template because the state requires this structure. Having all members on the “Board of Directors” creates a horizontal structure in keeping with the spirit of consensus.
 
The fourth, \textbf{Generic Organizational Structure}, is a model of how to arrange and operate your organization. Transparency is a value in consensus decisionmaking. Every organization should have a chart or outline that shows the kinds of meetings, committees and roles it utilizes and a handbook or document that explains the function and purpose of each meeting, committee or role.
